---
title: "RMBL survey marks for RTK GNSS"
output: github_document
---

```{r}
#| label: setup
#| include: FALSE

knitr::opts_chunk$set(echo = TRUE)
```

> DISCLAIMER: All data provided here, including coordinates of survey marks, are provided "as-is" and without any warranty of accuracy. The data and associated physical markers are shared exclusively for non-commercial research purposes to scientists at the Rocky Mountain Biological Laboratory and are categorically unfit for any other use (including, but not limited to, professional land surveying). Please refer to the [Colorado State Board of Licensure for Architects, Professional Engineers and Professional Land Surveyors](https://dpo.colorado.gov/AES) for information on professional land surveying and a directory of professional land surveyors licensed to conduct such work in Colorado.

## Background
Signals sent from GNSS satellites experience distortion as they pass through the atmosphere. Sub-foot and centimeter positioning with a GNSS receiver requires correcting for these distortions. This can be achieved in real-time (vs. post-processing) by sending corrections from a GNSS receiver placed at a known point (the "base" unit) to a second GNSS receiver that is moved around to collect data or re-find previously mapped positions (the "rover" unit). Because the base is stationary, any apparent changes in its measured position must be from atmospheric distortions. A nearby rover will experience a very similar set of atmospheric distortions, so it can correct its own position by using a constant stream of deviation measurements sent to it from the base, typically via an internet connection or a local radio signal.

Survey marks are most useful when position corrections are needed in real-time and internet connectivity is not available. Survey marks provide a stable point that can be re-found and reused repeatedly as the known location for a GNSS receiver to act as a base for nearby rovers. New marks established for real-time kinematic mapping should be established with care to ensure that the mark won't move, that its position can be measured with high precision, and that a base unit placed there can be "seen" by a rover unit across the entirety of the area to be mapped.

Emlid provides a brief overview and animation demonstrating how RTK works [here](https://docs.emlid.com/reachrs2/rtk-quickstart/rtk-introduction/).

## Maps of survey marks and estimated LoRa coverage
```{r}
#| label: maps
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
  
library(tidyverse)
library(sf)
library(ggrepel)
library(basemaps)

pts <- read_sf("current_mark-points.geojson") |>
  st_transform(crs = 3857)
vs <- read_sf("current_mark-viewsheds-8km.geojson") |>
  st_transform(crs = 3857) |>
  st_union()
bb <- st_buffer(pts, dist = 8000) |>
  st_bbox() |>
  st_as_sfc()
bb_gothic <- filter(pts, name %in% c("MexCut 3GPS", "Snodgrass TH",
                                     "East Maroon Pass")) |>
  st_buffer(dist = 2000) |>
  st_bbox() |>
  st_as_sfc()

city <- tribble(
  ~name, ~lat, ~lon,
  "Gunnison", 38.548056, -106.924722,
  "Crested Butte", 38.875278, -106.978056,
  # "Mt. Crested Butte", 38.908333, -106.960556,
  "Almont", 38.6647, -106.8462,
  "Pitkin", 38.608611, -106.515,
  "Sapinero", 38.459444, -107.302222,
  "Tincup", 38.754444, -106.478333,
  "Sargents", 38.404167, -106.415,
  "Powderhorn", 38.276944, -107.095833,
  "Schofield", 39.03748874575394, -107.06134674335976,
  "Pittsburg", 38.952222, -107.0625
) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(crs = 3857)

set_defaults(map_service = "esri", map_type = "world_hillshade")

ggplot()+
  basemap_gglayer(bb, verbose = FALSE)+
  geom_sf(data = vs, fill = "slateblue1", color = "transparent", alpha = 0.5)+
  geom_sf(data = bb_gothic, fill = "transparent", color = "black",
          linewidth = 0.75, linetype = "dashed")+
  geom_sf_text(data = city, aes(label = name), size = 3, color = "grey20")+
  geom_sf(data = pts, color = "slateblue")+
  geom_label_repel(data = pts, aes(label = name, geometry = geometry),
                   size = 2, stat = "sf_coordinates",
                   min.segment.length = 0, max.overlaps = 20, seed = 2029)+
  # geom_sf_label(data = bb_gothic, aes(label = "Inset"), size = 4.5,
  #              fontface = "bold", nudge_x = -20000)+
  scale_fill_identity() +
  ggtitle("Overview")+
  coord_sf(xlim = st_bbox(bb)[c("xmin", "xmax")],
           ylim = st_bbox(bb)[c("ymin", "ymax")],
           expand = FALSE)+
  theme(axis.title = element_blank())

ggplot()+
  basemap_gglayer(bb_gothic, verbose = FALSE)+
  geom_sf(data = vs, fill = "slateblue1", color = "transparent", alpha = 0.5)+
  geom_sf_text(data = city, aes(label = name), size = 3, color = "grey20")+
  geom_sf(data = st_filter(pts, bb_gothic), color = "slateblue")+
  geom_label_repel(data = st_filter(pts, bb_gothic), aes(label = name, geometry = geometry),
                   size = 2, stat = "sf_coordinates",
                   min.segment.length = 0, max.overlaps = 100)+
  scale_fill_identity() +
  ggtitle("Inset: Gothic corridor")+
  coord_sf(xlim = st_bbox(bb_gothic)[c("xmin", "xmax")],
           ylim = st_bbox(bb_gothic)[c("ymin", "ymax")],
           expand = FALSE)+
  theme(axis.title = element_blank())
```

## Choosing an existing survey mark for your research site
Four questions are useful when choosing a survey mark to use as a real-time base location:

1. How far is the proposed survey mark/base location from the mapping site?
The distance between a base unit and the rover (the "baseline") must be short enough to reliably send corrections over radio. Currently, LoRa is the most common between unit radio protocol. The transmission power is low and has a nominal range on the scale of a few kilometers (e.g., 8 km for Emlid RS2+). This range can be boosted with better antennas, signal repeaters, or different/more powerful radios (FCC license likely required!).

Additional problems arise with long baselines because the satellite signals received by the base and rover may experience different distortions if they pass through different parts of the atmosphere. The loss of precision will typically be on the order of a few parts per million. For example, an RTK setup with a 6 ppm loss of precision and a 12 km (7.5 mile) baseline will have an additional 72 mm (2.8 inch) error on top of the unit's nominal positional error. A heuristic limit is to keep the baseline <10 km (6.2 miles) for GNSS receivers that only use the L1 signal band and <60 km (37 miles) for L1/L2/+ multi-band receivers [(source)](https://docs.emlid.com/reachrs2/rtk-quickstart/rtk-introduction#single-band-and-multi-band-receivers), although the precision tolerances of your project may dictate more stringent limits.

2. Is there a clear line-of-sight between the survey mark and the entire area to be mapped?
Radio waves of the frequencies typically used for communication between GNSS receivers cannot pass through obstacles like dense forests and mountains very well--they are absorbed or they are reflected. Radio wave propagation in complex field environments is very challenging to predict. A useful heuristic is to use the line-of-sight rule: if the base and rover antennas can "see" each other and are within their nominal range, a radio signal should reliably pass between them. This repository contains [line-of-sight viewsheds](current_mark-viewsheds.geojson) for an 8 km radius around each survey mark. These represent *estimates* of where a pair of Emlid RS2 GNSS receivers could communicate using their out-of-the-box configuration. They are intended to aid in planning, and they should be validated in the field prior to operational use.

3. Is the survey mark suitable for the mapping application?
Survey marks in this database vary tremendously in...
- Accessibility from roads/trails: Accessibility can be a positive if it means that it's easier for your field crew to get a base receiver set up on the survey mark. Accessibility can become a liability if it means that you have to leave GNSS equipment unattended in a high traffic area where it is prone to disturbance from curious passersby or, worse, theft.

- Stability of the physical mark: Survey marks can be moved from their mapped location by physical processes like freeze/thaw cycles and slides of snow, earth, and rock. Information about the mark construction and most recent reported stability is provided in the database. Again, the survey mark to be used should be scouted for stability prior to operational use.

- Precision of the absolute mark position: There is not yet any consistent precision standard for mark inclusion in the database, nor have all marks been tied into the larger network of marks maintained by the [National Geodetic Survey](https://www.ngs.noaa.gov/datasheets/). Stable marks should be suitable for all relative positioning applications and meter accuracy absolute positioning.

## Suggested etiquette for mark use
Only one base GNSS receiver can occupy a given mark at a time, creating the potential for conflict if multiple users want to use the same base location. For RMBL-established marks, we suggest that you communicate with the Science Director about your use plan. Depending on equipment setups, it may be possible for multiple researchers to use the same correction stream from a single base. For marks established by an individual, we suggest contacting that individual to coordinate use of the mark. In most cases, the individual established the mark for their use in a particular location for which other existing marks were unsuitable, and it is included under their name on the Current Research Map. If you have set a mark, please ask that it be included on the Current Research Map under your name for as long as you are actively using it.

Conflicts can also arise if multiple bases are broadcasting RTK corrections on the same LoRa frequency (902.0-928.0 MHz). It's good practice to change the frequency your team uses to something different from the out-of-the-box default to minimize potential conflicts. The Science Director may have information on other researchers operating base stations near your research sites and may be able to help facilitate coordination between teams.

Longer term, there may be opportunities to set up a (semi)permanent base unit to broadcast corrections that cover a broad set of research sites with heavy demand for RTK. This will take a non-trivial amount of investment to set up and maintain. Alternatively, expansion of cellular data networks may allow wider use of new or existing web-based casters (e.g., [Mesa County RTVRN](https://www.mesacoumenty.us/departments-and-services/gis/gps-survey)).

## Releases and update schedule
Releases will be named following a semantic versioning scheme that follows the pattern `vYYYY.p`. `YYYY` corresponds to the calendar year and `p` is the patch number which starts at 0 and increments by 1. The previous major/minor release schedule is no longer necessary since the scripting of the viewshed workflow in v2024.3.

Update patches will be released on an as-needed basis, with the majority of activity occurring during the peak of field season activity (June-August). Updates will contain some combination of:

- survey mark and viewshed additions
- error corrections and improved mark position accuracy
- recovery/condition reports

The last patch for each calendar year will be submitted to the RMBL Current Research Map during the September-May "off-season."

## Submit a mark recovery attempt report
If you visit a survey mark, please consider submitting a [mark recovery attempt report](https://github.com/DICELab-NCSU/RMBL_survey_marks/issues/new?assignees=&labels=recovery-report&template=contribute-mark-recovery-attempt.md&title=Recovery+attempt%3A+%5Bmark+name%5D) (links to an Issues template for this repository). Survey marks can move or be damaged and destroyed, rendering them inaccurate or unusable. These reports will be used to update the condition columns associated with each survey mark.

Users are encouraged to collect high precision position data during mark recovery attempts if they are able to do so. Most of the marks were established using older GPS/GNSS technologies and/or a brief position recording duration. Collecting higher quality mark position data would enable a broader suite of research applications, especially as the RMBL Spatial Data Platform comes online. Please contact the maintainer of this database if you are interested in providing mark position data.

## Submit a record of a new mark
The procedure to submit a new mark is currently under development. Please contact the Science Director to arrange appropriate permissions if you are considering setting a new mark to aid your research. If you have set a mark for your research and would like to contribute it to the database, please open an 'Issue' at the top of this page.

## Technical documentation

### 1. Viewshed analysis

Viewsheds for each survey mark were estimated using the [`viewscape`](https://doi.org/10.32614/CRAN.package.viewscape) package v.`r packageVersion("viewscape")`.

#### 1.1. DEM dataset and pre-processing

Elevation data were sourced from the [USGS 3DEP 1/3rd arc-second Digital
Elevation
Models](https://www.sciencebase.gov/catalog/item/4f70aa9fe4b058caae3f8de5). The on-the-ground resolution is ~8m, which smooths out small terrain features that, strictly speaking, blocks line of sight but does not prevent local radio communication. Processing speeds are expected to be ~64× faster with this dataset compared to the highest available resolution DEM (~1m LiDAR).

The USGS 3DEP DEM files required are:

- USGS_13_n39w107_20220331.tif
- USGS_13_n39w108_20220720.tif
- USGS_13_n40w107_20220216.tif
- USGS_13_n40w108_20211208.tif

These files were merged, warped to EPSG:32613, and cropped to the [area
of interest](utilities/AOI.geojson) (also available as a Shapefile in
[utilities](utilities)). The AOI is defined as the bounding box that
encompasses the set of 10km-buffered survey marks.

The raster file for the DEM is not provided in this repo due to its large size, however
reproducible code is available in `code/2_merge-dem.R`.

#### 1.2. Calculating viewsheds

The code needed to reproduce the viewsheds is provided in `code/3_mk_viewsheds.R`. The 
observer locations are the [current_mark-points](current_mark-points.geojson), and we assume 
that the base and the rover are both on a 2m range pole or tripod. The stock LoRA antenna 
range for [Emlid RS2+](https://emlid.com/reachrs2/)) is 8km line-of-sight, but our testing 
with [aftermarket high gain antennas] (https://aeromao.com/product/triangl-extended-range-antenna-kit/) suggests that 11km LOS may be feasible under ideal conditions.

# Changelog

- **2025-05-26**: Update release schedule based on viewshed workflow changes in v2024.3.
- **2024-07-04**: Viewshed workflow was moved from QGIS to R for efficiency.
